### 2024-07-27 cluster-analysis.qmd
20240727 - cluster analysis

```{r}
library("mclust")
library("tidyverse")
```

@@

### Example {#sec-clusterAnalysisExample}

@@

tiersRB_boostrap <- mclustBootstrapLRT(
  player_stats_seasonal_offense_recentRB$fantasy_points,
  modelName = "V") # variable/unequal variance (for univariate data)

numTiersRB <- as.numeric(summary(tiersRB_boostrap)[,"Length"][1]) # or could specify the number of teams manually

tiersRB_boostrap
plot(
  tiersRB_boostrap,
  G = numTiersRB - 1)
```

@@

We can merge the player's classification into the dataset and plot each player's classification:

@@

@@

ggplot2::ggplot(
  data = player_stats_seasonal_offense_recentQB,
  mapping = aes(
    x = fantasy_points,
    y = position_rank,
    color = as.factor(tier)
  )) +
  geom_label(
    aes(
      x = fantasy_points,
      label = round(fantasy_points, 0),
      fill = as.factor(tier)
      ),
    color = "white",
    size = 5,
    #show.legend = FALSE
    ) +
  geom_label(
    aes(
      x = fantasy_points,
      label = player_display_name,
      fill = as.factor(tier)),
    color = "white",
    hjust = 1.1,
    size = 4,
    show.legend = FALSE
    ) +
  scale_y_continuous(trans = "reverse") +
  coord_cartesian(clip = "off") +
  theme_classic() +
  labs(
    x = "Projected Points",
    y = "Rank",
    title = "Quarterback Fantasy Points by Tier",
    fill = "Tier")
```

>>>>

```{r}
library("mclust")
library("tidyverse")
library("plotly")
```

@@

### Example: Tiers of Prior Season Fantasy Points {#sec-clusterAnalysisExample}

@@

numTiersRB <- 3

# The following code errors out:
#tiersRB_boostrap <- mclustBootstrapLRT(
#  player_stats_seasonal_offense_recentRB$fantasy_points,
#  modelName = "V") # variable/unequal variance (for univariate data)
#
#numTiersRB <- as.numeric(summary(tiersRB_boostrap)[,"Length"][1]) # or could specify the number of teams manually
#
#tiersRB_boostrap
#plot(
#  tiersRB_boostrap,
#  G = numTiersRB - 1)
```

@@

```{r}
clusterModelRBs <- Mclust(
  data = player_stats_seasonal_offense_recentRB$fantasy_points,
  G = numTiersRB,
)
```

```{r}
table(clusterModelRBs$classification)
```

```{r}
clusterModelWRs <- Mclust(
  data = player_stats_seasonal_offense_recentWR$fantasy_points,
  G = numTiersWR,
)
```

```{r}
table(clusterModelWRs$classification)
```

```{r}
clusterModelTEs <- Mclust(
  data = player_stats_seasonal_offense_recentTE$fantasy_points,
  G = numTiersTE,
)
```

```{r}
table(clusterModelTEs$classification)
```

#### Plot the Tiers {#sec-clusterAnalysisExamplePlotTiers}

We can merge the player's classification into the dataset and plot each player's classification.

##### Quarterbacks {#sec-clusterAnalysisExamplePlotTiersQB}

@@

player_stats_seasonal_offense_recentQB <- player_stats_seasonal_offense_recentQB %>%
  mutate(
    tier = factor(case_match( # recode so 1 is highest tier
      tier,
      1 ~ 4,
      2 ~ 3,
      3 ~ 2,
      4 ~ 1)))

@@

plot_qbTiers <- ggplot2::ggplot(
  data = player_stats_seasonal_offense_recentQB,
  mapping = aes(
    x = fantasy_points,
    y = position_rank,
    color = tier
  )) +
  geom_point(
    aes(
      text = player_display_name # add player name for mouse over tooltip
  )) +
  scale_y_continuous(trans = "reverse") +
  coord_cartesian(clip = "off") +
  labs(
    x = "Projected Points",
    y = "Position Rank",
    title = "Quarterback Fantasy Points by Tier",
    color = "Tier") +
  theme_classic() +
  theme(legend.position = "top")

ggplotly(plot_qbTiers)
```

##### Running Backs {#sec-clusterAnalysisExamplePlotTiersRBs}

```{r}
#| label: fig-rbTiers
#| fig-cap: "Running Back Fantasy Points by Tier."

player_stats_seasonal_offense_recentRB$tier <- clusterModelRBs$classification

player_stats_seasonal_offense_recentRB <- player_stats_seasonal_offense_recentRB %>%
  mutate(
    tier = factor(case_match( # recode so 1 is highest tier
      tier,
      1 ~ 3,
      2 ~ 2,
      3 ~ 1)))

player_stats_seasonal_offense_recentRB$position_rank <- rank(
  player_stats_seasonal_offense_recentRB$fantasy_points * -1,
  na.last = "keep",
  ties.method = "min")

plot_rbTiers <- ggplot2::ggplot(
  data = player_stats_seasonal_offense_recentRB,
  mapping = aes(
    x = fantasy_points,
    y = position_rank,
    color = tier
  )) +
  geom_point(
    aes(
      text = player_display_name # add player name for mouse over tooltip
  )) +
  scale_y_continuous(trans = "reverse") +
  coord_cartesian(clip = "off") +
  labs(
    x = "Projected Points",
    y = "Position Rank",
    title = "Running Back Fantasy Points by Tier",
    color = "Tier") +
  theme_classic() +
  theme(legend.position = "top")

ggplotly(plot_rbTiers)
```

##### Wide Receivers {#sec-clusterAnalysisExamplePlotTiersWRs}

```{r}
#| label: fig-wrTiers
#| fig-cap: "Quarterback Fantasy Points by Tier."

player_stats_seasonal_offense_recentWR$tier <- clusterModelWRs$classification

player_stats_seasonal_offense_recentWR <- player_stats_seasonal_offense_recentWR %>%
  mutate(
    tier = factor(case_match( # recode so 1 is highest tier
      tier,
      1 ~ 5,
      2 ~ 4,
      3 ~ 3,
      4 ~ 2,
      5 ~ 1)))

player_stats_seasonal_offense_recentWR$position_rank <- rank(
  player_stats_seasonal_offense_recentWR$fantasy_points * -1,
  na.last = "keep",
  ties.method = "min")

plot_wrTiers <- ggplot2::ggplot(
  data = player_stats_seasonal_offense_recentWR,
  mapping = aes(
    x = fantasy_points,
    y = position_rank,
    color = tier
  )) +
  geom_point(
    aes(
      text = player_display_name # add player name for mouse over tooltip
  )) +
  scale_y_continuous(trans = "reverse") +
  coord_cartesian(clip = "off") +
  labs(
    x = "Projected Points",
    y = "Position Rank",
    title = "Wide Receiver Fantasy Points by Tier",
    color = "Tier") +
  theme_classic() +
  theme(legend.position = "top")

ggplotly(plot_wrTiers)
```

##### Tight Ends {#sec-clusterAnalysisExamplePlotTiersTEs}

```{r}
#| label: fig-teTiers
#| fig-cap: "Tight End Fantasy Points by Tier."

player_stats_seasonal_offense_recentTE$tier <- clusterModelTEs$classification

player_stats_seasonal_offense_recentTE <- player_stats_seasonal_offense_recentTE %>%
  mutate(
    tier = factor(case_match( # recode so 1 is highest tier
      tier,
      1 ~ 4,
      2 ~ 3,
      3 ~ 2,
      4 ~ 1)))

player_stats_seasonal_offense_recentTE$position_rank <- rank(
  player_stats_seasonal_offense_recentTE$fantasy_points * -1,
  na.last = "keep",
  ties.method = "min")

plot_teTiers <- ggplot2::ggplot(
  data = player_stats_seasonal_offense_recentTE,
  mapping = aes(
    x = fantasy_points,
    y = position_rank,
    color = tier
  )) +
  geom_point(
    aes(
      text = player_display_name # add player name for mouse over tooltip
  )) +
  scale_y_continuous(trans = "reverse") +
  coord_cartesian(clip = "off") +
  labs(
    x = "Projected Points",
    y = "Position Rank",
    title = "Tight End Fantasy Points by Tier",
    color = "Tier") +
  theme_classic() +
  theme(legend.position = "top")

ggplotly(plot_teTiers)
```
